---
title: "CRISPR Screen and Gene Expression Differential Analysis"
author: "Lianbo Yu, Yue Zhao, and Lang Li"
date: "`r Sys.Date()`"
#output: rmarkdown::html_vignette
output: 
  pdf_document:
    toc: yes
    number_sections: true
vignette: >
  %\VignetteIndexEntry{CRISPR Screen and Gene Expression Differential Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction
CEDA is developed for analyzing read counts of single guide RNAs (sgRNAs)
by CRISPR screen experiments. sgRNAs are synthetically generated from genes 
and each gene can generate multiple sgRNAs. CEDA models sgRNA counts at 
different levels of gene expression by multi-component normal mixtures 
and EM algorithms. Posterior estimates at sgRNA level are then summarized 
for each gene.

In this document, we use a MDA231 cell experiment as an example to
demonstrate how to use CEDA to perform CRISPR screen data analysis.

## Data and Normalization
Three samples of MDA231 cells were untreated at T0, and another three 
samples of MDA231 cells were treated with DMSO at T0. We are interested 
in detecting sgRNAs that are differentially changed by treatment. 

sgRNA read counts along with a list of non-essential genes are stored 
in the dataset mda231 in CEDA. 
```{r data}
library(CEDA)
data("mda231")
dim(mda231$sgRNA)
length(mda231$neGene$Gene)
head(mda231$sgRNA)
```

sgRNA read counts needs to be normalized across sample replicates before 
formal analysis. Non-essential genes are assumed to have no change after 
DMSO treatment. Median normalization factors of the non-essential genes 
were used for normalizing sgRNA counts of all samples.
```{r normalization}
mda231.ne <- mda231$sgRNA[mda231$sgRNA$Gene %in% mda231$neGene$Gene,]
cols <- c(3:8)
mda231.norm <- medianNormalization(mda231$sgRNA[,cols], mda231.ne[,cols])[[2]]
```

## Analysis
Our goal is to detect essential sgRNAs that have different count levels 
between conditions. R package limma was used to calculate log fold ratios
between three untreated and three treated samples. 

### Calculating fold ratios
```{r limma}
library(limma)
group <- gl(2,3,labels=c("Control","Baseline"))
design <- model.matrix(~  0 + group)
colnames(design) <- sapply(colnames(design),function(x) substr(x,6,nchar(x)))
contrast.matrix <- makeContrasts("Control-Baseline",levels=design)
limma.fit <- limma(log2(mda231.norm+1),design,contrast.matrix)
```

Then results from limma analysis were merged with sgRNA counts.
```{r merge}
mda231.limma <- data.frame(mda231$sgRNA,limma.fit)
head(mda231.limma)
```

### Fold ratios under the null hypotheses
Under the null hypothses, all sgRNAs levels are unchanged between the
two conditions. To obtain fold ratios under the null, samples were 
permutated between two conditions, log fold ratios were obtained from
limma analysis under each permutation.
```{r betanull}
betanull <- limmaPermutation(log2(mda231.norm+1),design,contrast.matrix,20)
theta0 <- sd(betanull)
theta0
```

### Fitting three-component mixture models
A three-component mixture model is assumed for log fold ratios at 
different level of gene expression. Empirical Bayes method was employed 
to estimate parematers of the mixtures and posterior means were obtained 
for estimating actual log fold ratios between the two conditions. 
P-values of sgRNAs were then calculated by permutation method.

```{r mm, results='hide'}
nmm.fit <- normalMM(mda231.limma,theta0)
```

Results from the mixture model were shown in Figure $1$. False 
discovery rate of $0.05$ was used for declaring significant changes in red
color between the two conditions for sgRNAs.

```{r fig1, fig.cap = "Log fold ratios of sgRNAs vs. gene expression level"}
scatterPlot(nmm.fit$data,fdr=0.05,xlim(-0.5,12),ylim(-8,5))
```

### Gene level summarization
From the p-values of sgRNAs, gene level p-values were obtained by 
using modified robust rank aggregation method (alpha-RRA). Log 
fold ratios were also summarized at gene level.
```{r pval}
mda231.nmm <- nmm.fit[[1]]
p.gene <- calculateGenePval(exp(mda231.nmm$log_p), mda231.nmm$Gene, 0.05)
fdr.gene <- stats::p.adjust(p.gene$pvalue, method = "fdr")
lfc.gene <- calculateGeneLFC(mda231.nmm$lfc, mda231.nmm$Gene)
```

